A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-10 latex/main.tex
Updates from Overleaf

Although the exact location of a recombination breakpoint
is impossible to pinpoint in practice,
bounds can be placed by examining the parental and recombinant sequences.
% Cite VirusRecomb
In methods \cite{Alfonsi2024,Li2024CovRecomb}, and
approaches (e.g., Pango designation)
based on lineage-associated mutations,
the breakpoints are limited in precision to these mutations.
Similarly, methods based on probabilistic representations of
% Cite Bolotie
mutational profiles (e.g., Bolotie) suffer from this limitation.
% Does RIPPLES qualify as a genealogical method?
Genealogical methods, such as sc2ts and RIPPLES,
based on pinpointing exact parental \emph{sequences},
are in principle maximally precise,
assuming that the parental sequences
are well represented in the ARG.

@@

Using the size of a breakpoint interval as a proxy,
we quantified how precisely breakpoints could be
localized along the genome.
Across the QC-passing recombination events,
we found the median interval size to be 2,207 bases
(Figure~\ref{fig:recombinant_intervals}C),
indicating that the breakpoints of half of the events
could be narrowed to a small region (less than ~7\% of the genome),
while the uncertainty around the breakpoints of the other events
is considerably greater.
We further found a negative relationship between
breakpoint interval size and
the amount of divergence between recombinant parents,
which is measured by time to their MRCA
(Figure~\ref{fig:recombinant_intervals}D).
When the two parental lineages have diverged
for longer periods of time (the rightmost points),
more information becomes available
on either side of a breakpoint to pinpoint its location,
resulting in shorter intervals.
For instance, the breakpoint intervals tend to be the smallest
for events involving recombination between
the deeply divergent Omicron BA.1 and BA.2 sublineages
(the red points in Figure~\ref{fig:recombinant_intervals}D).

% Not interesting enough
Another key feature of genealogical methods
is that they can also provide \emph{time} intervals
reflecting uncertainty in the timing of recombination events.
We calculated the time interval of a recombination node
as the difference between the age of the youngest parent and
the age of the oldest child of the recombination node in the ARG
(either given as collection dates or estimated using tsdate).
We found that for three quarters of the QC-passing events,
the time interval is less than ~2 months
(Figure~\ref{fig:recombinant_intervals}D),
% Do we want to refer to the CIs around the node dates?
% Maybe we could use this result to lead to the next section?
% Connect to epidemiology a bit?
indicating that the ARG can precisely estimate
the emergence of recombination events.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination breakpoint intervals.
(A) Distribution of the breakpoint intervals of
354 recombination events along the genome
(annotated with gene labels on top).
Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) Distribution of breakpoint interval size (bases).
(D) Relationship between breakpoint interval size and
time between the recombination event and the parental MRCA.
The black curve shows the theoretical expected relationship
between these quantities,
estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
(E) Distribution of time interval size (months).
\label{fig:recombinant_intervals}
\end{figure}

>>>>

Although in practice
the exact location of a recombination breakpoint
is impossible to pinpoint,
bounds can be placed by
examining the parental and recombinant sequences.
In methods \cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb},
and approaches (e.g., Pango designation),
based on lineage-associated mutations,
the breakpoints are limited in precision to these mutations.
Similarly, methods based on probabilistic representations of
mutational profiles (e.g.\cite{Varabyou2021rapid})
suffer from this limitation.
However, genealogical methods, such as sc2ts and RIPPLES,
based on identifying exact parental \emph{sequences},
are in principle maximally precise,
assuming that the parental sequences are well represented in the ARG.

@@

Using the size of a breakpoint interval,
we measured how precisely breakpoints could
be localized along the genome.
Across the QC-passing recombination events,
we found the median breakpoint interval size to be 2,207 bases
(Figure~\ref{fig:recombinant_intervals}C),
indicating that the breakpoints of half of the events
could be narrowed to a small region (less than $\sim$7\% of the genome),
while the uncertainty around the breakpoints of the other events
is considerably large.
Further, we found a negative relationship between
the size of a breakpoint interval and
the amount of divergence between recombinant parents,
which is measured by time to their MRCA
(Figure~\ref{fig:recombinant_intervals}D).
When the two parental lineages have diverged
for longer periods of time (the rightmost points),
more information becomes available
on either side of a breakpoint to pinpoint its location,
resulting in shorter intervals.
For instance, the breakpoint intervals tend to be the smallest
for events involving recombination between
the deeply divergent Omicron BA.1 and BA.2 sublineages
(the red points in Figure~\ref{fig:recombinant_intervals}D).

Another key feature of genealogical methods
is that they can also provide \emph{time} intervals
which reflect uncertainty
in the estimated timing of recombination events.
The time interval of a recombination node in the ARG is bounded by
the age of the youngest parent and
the age of the oldest child of the recombination node
(either given as collection dates or estimated using tsdate).
We found that for three quarters of the QC-passing events,
the time interval spans less than $\sim$2 months
(Figure~\ref{fig:recombinant_intervals}D),
with long time intervals suggesting that
a parental lineage or the recombinant lineage
might be unsampled for an extended period of time.
% Do we want to refer to the CIs around the node dates?
This result indicates that the ARG can pinpoint
when new recombinant lineages emerge with fine resolution.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination breakpoint intervals.
(A) Distribution of the breakpoint intervals of
354 recombination events along the genome
(annotated with gene labels on top).
Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals per site.
(C) Distribution of breakpoint interval size (bases).
(D) Relationship between breakpoint interval size and
time between the recombination event and the parental MRCA.
The black curve shows the theoretical expected relationship
between these quantities,
estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).
(E) Distribution of time interval size (months).
}
\label{fig:recombinant_intervals}
\end{figure}
